---
title: "TCGA-SKCM data"
author: "Muhammet Emre KUŞ"
date: "07 April 2022"
output: html_document
---

# Summary of the dataset

__Project ID:TCGA-SKCM, 
dbGaP Study Accession:	phs000178,\
Name:	Skin Cutaneous Melanoma,\
Disease Type:	Nevi and Melanomas,\
Primary Site,	Skin,\
Program:	TCGA. 
[visit the website](https://portal.gdc.cancer.gov/projects/TCGA-SKCM)__

# For General Chunk Commands

```{r setup, include=TRUE}

knitr::opts_chunk$set(echo = TRUE)

```

#Load Libraries

```{r}

library("TCGAbiolinks")
library("limma")
library("edgeR")
library("glmnet")
library("factoextra")
library("FactoMineR")
library("caret")
library("SummarizedExperiment")
library("gplots")
library("survival")
library("survminer")
library("RColorBrewer")
library("gProfileR")
library("genefilter")

```

#Show the all TCGA projects

```{r}

# GDCprojects = getGDCprojects()
# 
# GDCprojects[c("project_id", "name")]

```

#Get details about the project

```{r}

TCGAbiolinks:::getProjectSummary("TCGA-SKCM")

```

#Querying RNA-seq data from project

```{r}

query <- GDCquery(
  
  project = "TCGA-SKCM",
                      
  data.category = "Transcriptome Profiling",
  
  workflow.type = "STAR - Counts",
  
  data.type = "Gene Expression Quantification")

```

#Querying Clinical Metadata from project

```{r}

clin <- GDCquery_clinic("TCGA-SKCM", type = "clinical", save.csv = F)

```

#Visualize the results in a better way 

```{r}

lihc_res = getResults(query)

colnames(lihc_res)

dim(lihc_res)

```

#Show some spesific information

```{r}

head(lihc_res$sample_type)

lihc_res$sample_type1<-factor(lihc_res$sample_type,levels=c("Metastatic","Primary Tumor","Additional Metastatic","Solid Tissue Normal"))

summary(lihc_res$sample_type1)

```

#GDC Download

```{r}

GDCdownload(
  query,
  method = "api",
  files.per.chunk = 10
)

```

#Load the actual RNASeq

```{r}

tcga_data <- GDCprepare(query)

dim(tcga_data)

head(colnames(colData(tcga_data)))

tail(rownames(rowData(tcga_data)))

assays(tcga_data)

```

#Exploring the Data

```{r}

table(tcga_data@colData$vital_status)

table(tcga_data@colData$age_at_index)

table(tcga_data@colData$gender)

```

#Gene Expression Matrices

```{r}

dim(assay(tcga_data))

head(assay(tcga_data)[,1:10])

head(assay(tcga_data)[,1:10])

```

#Save the Data

```{r}

saveRDS(object = tcga_data,
        file = "tcga_data.RDS",
        compress = FALSE)

```

#Load the Data 

```{r}

tcga_data = readRDS(file = "tcga_data.RDS")

```

#RNASeq Normalization

```{r}

limma_pipeline = function(
  tcga_data,
  condition_variable,
  reference_group=NULL){

  design_factor = colData(tcga_data)[, condition_variable, drop=T]

  group = factor(design_factor)
  if(!is.null(reference_group)){group = relevel(group, ref=reference_group)}

  design = model.matrix(~ group)

  dge = DGEList(counts=assay(tcga_data),
                 samples=colData(tcga_data),
                 genes=as.data.frame(rowData(tcga_data)))

  # filtering
  keep = filterByExpr(dge,design)
  dge = dge[keep,,keep.lib.sizes=FALSE]
  rm(keep)

  # Normalization (TMM followed by voom)
  dge = calcNormFactors(dge)
  v = voom(dge, design, plot=TRUE)

  # Fit model to data given design
  fit = lmFit(v, design)
  fit = eBayes(fit)

  # Show top genes
  topGenes = topTable(fit, coef=ncol(design), number=100, sort.by="p")

  return(
    list(
      voomObj=v, # normalized data
      fit=fit, # fitted model and statistics
      topGenes=topGenes # the 100 most differentially expressed genes
    )
  )
}

```

#classification

```{r}

limma_res = limma_pipeline(
  tcga_data=tcga_data,
  condition_variable="sample_type",
  reference_group="Metastatic"
)

```

#Save the object

```{r}

saveRDS(object = limma_res,
        file = "limma_res.RDS",
        compress = FALSE)

```

#grouping patients by gender

```{r}

gender_limma_res = limma_pipeline(
  tcga_data=tcga_data,
  condition_variable="gender",
  reference_group="female"
)

```

#use as class for grouping the samples

```{r}

clinical_data = colData(tcga_data)

group = factor(clinical_data$definition)

group = relevel(group, ref="Metastatic")

summary(group)
```

#Create design matrix

```{r}

design = model.matrix(~group)
head(design)

```

#remove genes, which have low amount of counts

```{r}

dge = DGEList( # creating a DGEList object
  counts=assay(tcga_data),
  samples=colData(tcga_data),
  genes=as.data.frame(rowData(tcga_data)))

# filtering
keep = filterByExpr(dge,design) # defining which genes to keep

dge = dge[keep,,keep.lib.sizes=FALSE] # filtering the dge object

rm(keep) #  use rm() to remove objects from memory if you don't need them anymore

```

#voom

```{r}

dge = calcNormFactors(dge,method="TMM")

v = voom(dge,design,plot=TRUE)

fit = lmFit(v, design)

fit = eBayes(fit)

topGenes = topTable(fit, coef=1, sort.by="p")

print(topGenes)

```

#sessioninfo

```{r}

sessionInfo()

```

